---
output:
  github_document:
    html_preview: false
bibliography: references.bib
link-citations: true
---

<!-- README.md is generated from README.Rmd. Please edit that file, then run rmarkdown::render("README.Rmd"). -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sliR

<!-- badges: start -->
[![R-CMD-check](https://github.com/LezzGitIt/sliR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/LezzGitIt/sliR/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

sliR computes the **Standardized Length Index (SLI)**, a size correction for
appendage lengths, and simulates allometric morphological data with a known
scaling exponent, correlation structure, and error.

The SLI adapts the scaled mass index of @peigNewPerspectivesEstimating2009.
Where their index standardises *mass* to a reference *length*, the SLI
standardises a *length* to a reference *mass*:

$$SLI_i = A_i \left( \frac{M_0}{M_i} \right)^{b}$$

for individual $i$ with appendage length $A_i$ and body mass $M_i$, an arbitrary
reference mass $M_0$, and an allometric scaling exponent $b$. The allometric
scaling exponent is calculated using standardized major axis (SMA) regression,
as recommended by @wartonSmatr3Package2012.

## Installation

```r
# install.packages("remotes")
remotes::install_github("LezzGitIt/sliR")
```

## Computing the SLI

Simulate a set of individuals, then size-correct their appendage lengths under
isometric scaling ($b = 1/3$, so linear dimensions scale with the cube root
of mass).

```{r sli-basic}
library(sliR)

set.seed(1)
birds <- sim_allometric(n = 500, b_sma = 1/3, r_app_mass = 0.3)

calc_sli(birds, b_sli = 0.33)
```

`calc_sli()` appends a column and returns the rows in their original order. The function
can be piped, creating several indices that sit side by side for comparison.

```{r sli-chain, message = FALSE}
library(dplyr)

est_b <- coef(smatr::sma(Append_log ~ Mass_log, data = birds))[["slope"]]

birds |>
  calc_sli(b_sli = 0.33,  rename_col = "sli_isometry") |>
  calc_sli(b_sli = est_b, rename_col = "sli_estimated") |>
  select(Append, Mass, sli_isometry, sli_estimated)
```

$M_0$ is an arbitrary reference; changing it rescales every SLI by a constant. 
The SLI maintains the units of the appendage $A$ and projects it along the line 
implied by the scaling coefficient to the reference mass $M_0$. The ranking of 
individuals will never change by changing $M_0$. $M_0$ defaults to the mean of `Mass`.

If your columns are not named `Append` and `Mass`, you can name them yourself — the
arguments are unquoted:

```{r sli-nse}
lowercase <- rename(birds, wing = Append, mass = Mass)
calc_sli(lowercase, Append = wing, Mass = mass, b_sli = 0.33)
```

## Group-specific allometric slopes

A single exponent for the whole sample assumes every group scales alike. When
age or sex classes scale differently, pass them to `control` and each
individual is corrected with their own group's SMA slope.

```{r slopes}
birds$Sex <- rep(c("F", "M"), length.out = nrow(birds))
birds$Age <- rep(c("Juv", "Adult"), each = 2, length.out = nrow(birds))

build_sli_slopes_tbl(birds, control = c("Age", "Sex"))
```

Note the structure: with two control variables this fits *two* models — mass
interacted with age, and mass interacted with sex — and averages their slopes
per cell, rather than fitting a single age × sex interaction. That keeps each
slope estimated at the marginal rather than the joint sample size.

Averaging hides disagreement, so when a cell's per-variable slopes differ by more
than `slope_diff_warn` (0.15 by default) you get a warning naming the cell with the largest discrepancy.
Inspect the `b_sma_*` columns before trusting `b_sli_avg`.

`calc_sli()` calls this for you. Individuals whose group is `NA` or unknown-coded
receive `sli = NA` and keep their row.

```{r sli-control}
calc_sli(birds, control = c("Age", "Sex")) |>
  select(Append, Mass, Age, Sex, sli)
```

Before trusting per-group slopes, check that each group *has* an allometric
relationship to fit. A group whose mass and appendage length are uncorrelated
yields a SMA slope that is noise.

```{r cor-tbl}
build_group_cor_tbl(birds, control = c("Age", "Sex"))
```

## Simulating allometric data

`sim_allometric()` draws from a bivariate log-normal specified by its allometry
rather than by raw standard deviations. That allometry has only **two degrees of
freedom**: any two of the correlation `r_app_mass` and the three slopes `b_ols`,
`b_sma`, `b_avg` determine the other two. Supply two, or none to accept the
isometric default `b_sma = 1/3, r_app_mass = 0.3`.

The three slopes are not interchangeable. `b_ols` is the slope of
$E[\log A \mid \log M]$, and so the *generative* slope an `lm()` recovers.
`b_sma` is $\sigma_A / \sigma_M$, the slope an SMA fit returns and the exponent
`calc_sli()` assumes. Note that there is no clear consensus regarding which line-fitting technique is best for allometric scaling and related applications. Thus, `b_avg` chooses the midpoint between  `b_ols` and  `b_sma`, which no estimator returns; it is a convention for splitting the difference. `implied_allometry()` shows what any parameter set implies:

```{r implied}
implied_allometry(b_sma = 1/3, r_app_mass = 0.3)
```

```{r sim-check}
set.seed(3)
d <- sim_allometric(n = 2000, b_sma = 1/3, r_app_mass = 0.3, trim_sd = NULL)

c(b_ols = coef(lm(Append_log ~ Mass_log, data = d))[[2]],
  b_sma = coef(smatr::sma(Append_log ~ Mass_log, data = d))[["slope"]])
```

### Simulating along an environmental gradient

Pass `gradient` to add an environmental variable that both traits respond to,
and give its correlation with each log trait. **The gradient is whatever you
name it** — `"Temperature"`, `"Latitude"`, `"Year"`, `"Rainfall"`, `"Elevation"`,
`"Aridity"` — the string simply becomes the column name. `gradient_range` then
gives the units you want it expressed in. Omit `gradient_range` and you get a
z-scored gradient (mean 0, SD 1).

Because a gradient is a *sampling design* variable rather than a third trait, it
is drawn uniformly by default, so `gradient_range` is honoured exactly and every
part of the gradient is sampled equally.

The gradient and allometry arguments compose freely: the allometry describes how
appendage scales with mass, the gradient describes how each responds to the
environment.

```{r gradient}
set.seed(4)
gulls <- sim_allometric(
  n = 2000,
  # allometry: a shallower-than-isometric SMA slope, tightly correlated
  b_sma = 0.28, r_app_mass = 0.6, sd_log_mass = 0.09,
  # gradient: longer wings and heavier birds toward the poles
  gradient = "Latitude", gradient_range = c(-90, 90),
  r_grad_app = -0.3, r_grad_mass = 0.2,
  trim_sd = NULL
)

range(gulls$Latitude)
c(r_grad_app = cor(gulls$Latitude, gulls$Append_log),
  b_sma      = coef(smatr::sma(Append_log ~ Mass_log, data = gulls))[["slope"]])
```

A normal gradient (`gradient_dist = "normal"`) reproduces the usual
three-variable multivariate normal, but is unbounded — it will happily emit
latitudes beyond $\pm 90$ — and leaves the extremes of the gradient nearly
unsampled.

Not every set of correlations is attainable. An appendage that lengthens while
mass falls along the gradient is incompatible with a strong positive
appendage–mass correlation, and `sim_allometric()` says so rather than quietly
approximating.

Note also that a correlation structure is **not** a causal structure. A gradient
correlated with both traits cannot distinguish a direct effect on the appendage
from an indirect one acting through body size.

### Measurement and biological error

Error comes in two flavours, each a fraction of the trait's own standard
deviation. `meas_error` is observer error and hits both traits; `transient_error_*`
is genuine short-term biological fluctuation, which afflicts mass far more than
a skeletal appendage. Because that fluctuation is independent of the appendage,
it attenuates the observed relationship:

```{r sim-error}
set.seed(5)
clean <- sim_correlated(n = 2000, r_app_mass = 0.3)
set.seed(5)
noisy <- sim_correlated(n = 2000, r_app_mass = 0.3, transient_error_mass = 1)

c(clean = cor(clean$Append, clean$Mass),
  noisy = cor(noisy$Append, noisy$Mass))
```

`sim_correlated()` is the simpler raw-scale counterpart, with no log transform
and no allometric target — useful for building intuition about how a fitted
slope responds to correlation and to error on one trait but not the other.
`build_cov_mat()` exposes the covariance matrix that `sim_allometric()` consumes.

## Function reference

| Function | Purpose |
|---|---|
| `calc_sli()` | Standardized Length Index per individual |
| `build_sli_slopes_tbl()` | Per-group SMA allometric slopes |
| `build_group_cor_tbl()` | Per-group diagnostic for whether allometry is meaningful |
| `sim_allometric()` | Log-normal traits with a target allometry, optionally along a gradient |
| `implied_allometry()` | What a set of allometric parameters implies for the remaining parameters |
| `sim_correlated()` | Raw-scale correlated appendage and mass |
| `sim_grid()` | Map `sim_allometric()` over a grid of parameter combinations |
| `build_cov_mat()` | Log-scale covariance matrix behind `sim_allometric()` |

## References

<div id="refs"></div>
