---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file, then run rmarkdown::render("README.Rmd"). -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sliR

<!-- badges: start -->
<!-- badges: end -->

sliR computes the **Standardized Length Index (SLI)**, a size correction for
appendage lengths, and simulates allometric morphological data with a known
scaling exponent, correlation structure, and error.

The SLI adapts the scaled mass index of [Peig & Green
(2009)](https://doi.org/10.1111/j.1600-0706.2009.17643.x). Where their index
standardises *mass* to a reference *length*, the SLI standardises a *length* to
a reference *mass*:

$$SLI_i = A_i \left( \frac{L_0}{M_i} \right)^{b}$$

for individual $i$ with appendage length $A_i$ and body mass $M_i$, an arbitrary
reference mass $L_0$, and an allometric scaling exponent $b$.

## Installation

```r
# install.packages("remotes")
remotes::install_github("LezzGitIt/sliR")
```

## Computing the SLI

Simulate a set of individuals, then size-correct their appendage lengths under
geometric similarity ($b = 1/3$, so linear dimensions scale with the cube root
of mass).

```{r sli-basic}
library(sliR)

set.seed(1)
birds <- sim_allometric(n = 500, b_avg = 0.33, r_am = 0.3)

calc_sli(birds, b_sli = 0.33)
```

`calc_sli()` appends a column and returns the rows in their original order, so
it chains: several indices can sit side by side for comparison.

```{r sli-chain, message = FALSE}
library(dplyr)

est_b <- coef(smatr::sma(Append_log ~ Mass_log, data = birds))[["slope"]]

birds |>
  calc_sli(b_sli = 0.33,  rename_col = "sli_isometry") |>
  calc_sli(b_sli = est_b, rename_col = "sli_estimated") |>
  select(Append, Mass, sli_isometry, sli_estimated)
```

$L_0$ is an arbitrary reference: changing it rescales every SLI by a constant,
which changes the index's units but never the ranking of individuals or the
result of any subsequent analysis. It defaults to the mean of `Mass`.

If your columns are not named `Append` and `Mass`, name them yourself — the
arguments are unquoted:

```{r sli-nse}
lowercase <- rename(birds, wing = Append, mass = Mass)
calc_sli(lowercase, Append = wing, Mass = mass, b_sli = 0.33)
```

## Group-specific allometric slopes

A single exponent for the whole sample assumes every group scales alike. When
age or sex classes have their own allometry, pass them as `control` and each
individual is corrected with their own group's SMA slope.

```{r slopes}
birds$Sex <- rep(c("F", "M"), length.out = nrow(birds))
birds$Age <- rep(c("Juv", "Adult"), each = 2, length.out = nrow(birds))

build_sli_slopes_tbl(birds, control = c("Age", "Sex"))
```

Note the structure: with two control variables this fits *two* models — mass
interacted with age, and mass interacted with sex — and averages their slopes
per cell, rather than fitting a single age × sex interaction. That keeps each
slope estimated at the marginal rather than the joint sample size.

`calc_sli()` calls this for you. Individuals whose group is `NA` or unknown-coded
receive `sli = NA` and keep their row.

```{r sli-control}
calc_sli(birds, control = c("Age", "Sex")) |>
  select(Append, Mass, Age, Sex, sli)
```

Before trusting per-group slopes, check that each group *has* an allometric
relationship to fit. A group whose mass and appendage length are uncorrelated
yields a slope that is noise.

```{r cor-tbl}
build_group_cor_tbl(birds, control = c("Age", "Sex"))
```

## Simulating allometric data

`sim_allometric()` draws from a multivariate log-normal whose covariance is
built from a *target allometric slope* rather than from raw standard deviations.
`b_avg` is the mean of the OLS and SMA slopes of log appendage on log mass:

```{r sim-check}
set.seed(3)
d <- sim_allometric(n = 2000, b_avg = 0.33, r_am = 0.3, trim_sd = NULL)

b_ols <- coef(lm(Append_log ~ Mass_log, data = d))[[2]]
b_sma <- coef(smatr::sma(Append_log ~ Mass_log, data = d))[["slope"]]

c(b_ols = b_ols, b_sma = b_sma, b_avg = mean(c(b_ols, b_sma)))
```

Error comes in two flavours, each a fraction of the trait's own standard
deviation. `meas_error` is observer error and hits both traits; `transient_error_*`
is genuine short-term biological fluctuation, which afflicts mass far more than
a skeletal appendage. Because that fluctuation is independent of the appendage,
it attenuates the observed relationship:

```{r sim-error}
set.seed(5)
clean <- sim_correlated(n = 2000, r = 0.3)
set.seed(5)
noisy <- sim_correlated(n = 2000, r = 0.3, transient_error_mass = 1)

c(clean = cor(clean$Append, clean$Mass),
  noisy = cor(noisy$Append, noisy$Mass))
```

`sim_correlated()` is the simpler raw-scale counterpart, with no log transform
and no allometric target — useful for building intuition about how a fitted
slope responds to correlation and to error on one trait but not the other.
`build_cov_mat()` exposes the covariance matrix that `sim_allometric()` consumes.

## Function reference

| Function | Purpose |
|---|---|
| `calc_sli()` | Standardized Length Index per individual |
| `calc_lambda()` | Ratio of squared CVs; its square root approximates the SMA slope |
| `build_sli_slopes_tbl()` | Per-group SMA allometric slopes |
| `build_group_cor_tbl()` | Per-group diagnostic for whether allometry is meaningful |
| `sim_allometric()` | Log-normal traits with a target allometric slope |
| `sim_correlated()` | Raw-scale correlated appendage and mass |
| `build_cov_mat()` | Log-scale covariance matrix behind `sim_allometric()` |

## Reference

Peig, J. & Green, A.J. (2009) New perspectives for estimating body condition
from mass/length data: the scaled mass index as an alternative method. *Oikos*
118, 1883–1891.
