---
title: "Getting started with sliR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with sliR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(sliR)
library(dplyr)
library(purrr)
```

This vignette covers the two things a user of sliR actually does: simulating
data across a grid of parameters, and computing the SLI on real data with
messy, multi-level grouping.

## Simulating across a parameter grid

`sim_allometric()` draws one dataset per call. To compare many scenarios,
build a tibble of parameter combinations and map over its rows with
`purrr::pmap()` — this works because the grid's column names match
`sim_allometric()`'s arguments exactly.

```{r grid-no-gradient}
grid <- tibble::as_tibble(expand.grid(
  n = 300,
  r_app_mass = c(0.1, 0.3, 0.5),
  b_sma = c(1 / 4, 1 / 3)
))
grid

set.seed(1)
sims <- pmap(grid, sim_allometric)
map_dbl(sims, \(d) cor(d$Append_log, d$Mass_log))
```

Each element of `sims` is one simulated dataset for one row of `grid`.

A gradient (temperature, latitude, year, ...) needs two extra columns whose
names are fixed by `sim_allometric()`: `r_grad_app` and `r_grad_mass`. The
gradient's name itself (`gradient = "Temperature"`) is not part of the grid,
because every row shares it — pass it as a separate, constant argument to
`pmap()`.

```{r grid-gradient}
grad_grid <- tibble::as_tibble(expand.grid(
  n = 300,
  r_grad_app = c(-0.3, 0, 0.3),
  r_grad_mass = -0.1
))
grad_grid

set.seed(1)
grad_sims <- pmap(
  grad_grid, sim_allometric,
  gradient = "Temperature", gradient_range = c(0, 24)
)
map_dbl(grad_sims, \(d) cor(d$Append_log, d$Temperature))
```

## Per-group SLI on messy real data

Real datasets rarely have every individual assigned a clean group: some may
be age-unknown, and some age classes may be too sparse — or too weakly
allometric — for a per-group slope to mean anything. Simulate something in
that spirit: two age classes with a real mass–appendage relationship, one
(juveniles) without, plus some unknown-coded individuals.

```{r messy-data}
set.seed(1)
adult <- sim_allometric(n = 150, r_app_mass = 0.5, b_sma = 1 / 3) |> mutate(Age = "Adult")
subadult <- sim_allometric(n = 150, r_app_mass = 0.45, b_sma = 1 / 3) |> mutate(Age = "Subadult")
juvenile <- sim_allometric(n = 150, r_app_mass = 0.02, b_sma = 1 / 3) |> mutate(Age = "Juvenile")
unk <- sim_allometric(n = 10) |> mutate(Age = "Unk")

birds <- bind_rows(adult, subadult, juvenile, unk)
count(birds, Age)
```

Before trusting a per-group slope, check whether each group's mass and
appendage length are actually correlated with `build_group_cor_tbl()`. Rows
coded `"Unk"` (the default in `unknown_codes`) are excluded automatically, so
they never appear here.

```{r group-cor}
build_group_cor_tbl(birds, control = "Age")
```

`Age == "Juvenile"` has no meaningful mass–appendage relationship (`r` near
zero, large `p_value`), so an SMA slope fit to that group would be noise.
`calc_sli()` fits a slope to every level present in the data, with no
significance filtering of its own, so drop the juveniles before calling it —
the remaining unknown-coded rows still come along, and still receive
`sli = NA` automatically.

```{r calc-sli}
birds |>
  filter(Age != "Juvenile") |>
  calc_sli(control = "Age") |>
  select(Append, Mass, Age, sli) |>
  filter(is.na(sli) | Age == "Adult") |>
  head()
```

Individuals whose `control` value is `NA` (here, the `"M"` birds we recoded,
plus the original `"Unk"` rows once included) receive `sli = NA` rather than
a slope fit to a relationship that isn't there.
